Characterising cheque dishonour cases in India: Causes for delays and policy implications

### 2022-04-18 DOC/introduction.tex
Editing the main paper. Added the marked points

\begin{description}
\item[H$_{1a}$]: cases, where the accused does not appear for one or more hearings, require more days to dispose
\item[H$_{1b}$]: cases, where the accused does not appear for one or more hearings, require more hearings to dispose
\end{description}

@@

\begin{description}
\item[H$_{2a}$]: cases converted to summons trials require more days to dispose
\item[H$_{2b}$]: cases converted to summons trials require more hearings to dispose
\end{description}

@@

\begin{description}
\item[H$_{3a}$]: cases referred to mediation require more days to dispose
\item[H$_{3b}$]: cases referred to mediation require more hearings to dispose
\end{description}

>>>>

\begin{description}
\item[H$_{1a}$]: Cases, where the accused does not appear for one or more hearings, require more days to dispose
\item[H$_{1b}$]: Cases, where the accused does not appear for one or more hearings, require more hearings to dispose
\end{description}

@@

\begin{description}
\item[H$_{2a}$]: Cases converted to summons trials require more days to dispose
\item[H$_{2b}$]: Cases converted to summons trials require more hearings to dispose
\end{description}

@@

\begin{description}
\item[H$_{3a}$]: Cases referred to mediation require more days to dispose
\item[H$_{3b}$]: Cases referred to mediation require more hearings to dispose
\end{description}

### 2022-04-18 DOC/paper_chequeDishonour_combined.tex
Editing the main paper. Added the marked points

\graphicspath{{../DOC/SLIDES/IMAGES/}}

@@

India has a slow judiciary - courts are clogged with large backlogs \autocite{moog1992delays, debroy2008justice, dutta2019modernise}. As of 8th December 2021, over forty-one million cases were pending across district courts \autocite{njdg2021}. It is estimated that judicial delays cost India around 1.5\% of its GDP annually \autocite{dey2016_cost}. Slow judiciaries also have adverse consequences on the structure and efficiency of markets and the quality of life of citizens \autocite{world2004world, chemin2007impact, rao2020institutional}. Therefore, it is hardly surprising that tackling judicial delay has become a top priority for Indian judges and policymakers.

@@

Given this increasing volume and persistent mention of \gls{ni} cases as a major cause of judicial delay, \cite{sridhar2017_cheque} examined the behaviour of cheque dishonour cases in India by looking at the DAKSH database of 67,000 cases spread across Indian courts. In most cases, she found that resolution is delayed well beyond the statutorily prescribed timeline of 180 days. Using 50,000 case level data for eight States and two Union Territories, we also estimate the survival probability (using non-parametric Kaplan-Meier estimation) to understand the proportion of unresolved cases within the stipulated time.\footnote{The dataset used for this analysis is documented in section \ref{sec:data-description}.} Figure \ref{fig:stateSurvival} shows the survival probability (case ongoing) for the \gls{ni} cases for different States. As per our estimate, less than 20\% of the cases get disposed of within the prescribed timeline of 180 days. Also, the probability of case completion within 500 days is less than 50\% for most States.

\begin{figure}[!ht]
\centering
\caption{Survival probability of cases across States}\label{fig:stateSurvival}
\footnotesize
\textit{Note:} The X-axis shows the number of days that a case takes from the date of admission until the outcome. The Y-axis shows the probability of the case continuing up to a given number of days.
\includegraphics[scale=0.35]{survstates.pdf}
\end{figure}

Since cheque dishonour cases are a large burden on the judiciary, various committees, institutions, and government bodies have focused on expediting disposal. For example, the \cite{lci2008_138}, by relying on newspaper reports concerning the proportion of cheque dishonour cases, recommended setting up Fast Track Magisterial Courts (for a detailed discussion, refer to \cite{bhan2015_placing}). The Act was also subject to amendments in 2015 and 2018 to deal with the various procedural matters. Similarly, the Supreme Court has issued directions to lower courts and various financial institutions to take a more pragmatic and realistic approach for the speedy disposal of \gls{ni} cases.

In 2020, the Supreme Court of India took on board a suo-motu case concerning the “expeditious trial of cases under section 138 of the Negotiable Instruments Act 1881” \autocite{sc2020_138}. Among other directions, the court formed a \gls{coe}\footnote{Headed by Hon’ble Mr. Justice RC Chavan, former Judge of the Bombay High Court.} and appointed Amici Curiae\footnote{Mr. Sidharth Luthra (Sr Advocate) and Mr. K Parameshwar (Advocate).} to assist the court, to study processes to expedite the disposal of complaints under \S~138 of the \gls{ni}. In late 2020, the Amici submitted their recommendations. This included (i) increasing the use of pre-and post-summons mediation, (ii) expediting the service of summons to reduce absconsion, (iii) addressing the multiplicity of proceedings, etc. (for additional detail, refer to \cite{amicus2020_submission}).

Given this background, the purpose of this paper is fourfold. First, using case level eCourts data, for the first time in India, we estimate the real judicial caseload due to the \gls{ni}. Second, we estimate the average case disposal time and its variation over time and across States. Third, we build a conceptual framework to establish the link between specific case characteristics and procedural delay. Furthermore, we empirically test our conceptual framework of procedural determinants of case disposal time using regression analysis. In line with \cite{bielen2015}, using case level interim and final orders data, we establish the link between the specific case characteristics and procedural issues while controlling for inherent court and judge characteristics. The empirical literature on the length of court proceedings currently focuses mainly on determinants at the country and court levels, e.g. public budgets for courts, backlogs, number of judges, etc. Some studies also focus on the case level determinants like the number of parties involved, number of pleadings, etc. Given access to the interim and final orders data, we focus on the case characteristics that can be directly targeted using procedural reforms.

@@

Our estimates suggest that cheque dishonour cases represent 13.2\% of courts' workload (pending and disposed) in India. This number could be underestimated because, in about 36\% of the cases in the sample, orders were either unavailable or could not be parsed. We also find significant variation in case disposal time across states. Median disposal time varies from as low as 218 days in Karnataka to as high as 547 days in Himachal Pradesh, with a median disposal time of 395 days. We also find that only a fifth of the cases gets disposed of within the prescribed limit of six months.

Controlling for State and year fixed effects, our findings suggest that, all else being equal, cases where the accused fails to appear take an additional 201 days and 7 hearings to dispose of. Similarly, cases run as summons trials take 112 days and 7.4 hearings more than summary trials. Cases with jurisdictional issues take 271 days and 5.5 hearings more to dispose of than cases without these issues. Cases with a multiplicity of proceedings take an additional 168 days and 9.9 hearings to dispose of than those without. Contested cases take fewer days to dispose of than uncontested cases, even though they take more hearings. Our results also indicate that the 2015 amendments have significantly reduced case duration -- reducing the case duration by more than six months.

@@

\begin{enumerate}
\item how to bring the accused to the court?;
\item how to run the trials (summary or summons trials)?;
\item whether to refer the case to mediation;
\item how to clarify jurisdictional issues?; and
\item how to avoid the multiplicity of proceedings?
\end{enumerate}

@@

We first drew a random sample of 100,000 cases filed between 1st January 2014 and 31st December 2018 across India from the database published by Development Data Lab.\footcite{devdatalabs2021_eCourtsData} {\color{red} Devendra: Briefly describe DevDataLab data in two lines and why we are using this.} This window was chosen based on consultations with other researchers and practising advocates on the quality and completeness of data. We then attempted to download the case details for these 100,000 cases from the eCourts database.\footcite{ecourts2022} We were able to download case details for approximately 89,000 cases.\footnote{This difference was caused by non-functional links and corrupted or non-machine data.} We used regular expressions to check whether the ``Act Name'' field references the \gls{ni}. This was the first level of case classification. However, from our previous experience with eCourts data, we expected NI Act cases among those where the Act Name field did not match the search parameters.\footcite[The Act Name field sometimes does not contain the name of the substantive Act but rather the name of the procedural Act (e.g., CrPC) or, in rare cases, the subject matter of the case. This is mainly because case classification and nomenclature practices vary from State to State. However, in rare cases, the field is not populated or contains erroneous information. For details, please see,][]{damle2020_ecourtsData} So we used regular expressions to search for references to the \gls{ni}, cheque bounce/dishonour along with \S~138 in the interim and final order texts, where available. We tagged the cases found through this protocol as related to cheque dishonour under \S~138 of the \gls{ni}.

@@

\item [Contested cases] -- {\color{red} Devendra: How to figure out whether the case is contested or not? Add text here}

Table \ref{tab:cont_yearWise} shows the number of cases in the sample which were contested by the accused. The proportion of such cases is significant but appears to have declined over the years. This may be due to an increased tendency of parties to compound a case.\footnote{\S~147 of the Negotiable Instruments Act provides that every offence punishable under the Act shall be compoundable. \S~147, Negotiable Instruments Act. Offences to be compoundable.} We expected contested cases to take more days and more hearings to dispose of than uncontested cases.

@@

\begin{table}[!ht]
\caption{Cases with identified characteristics of interest}
\label{tab:case_chars}
\centering
\footnotesize
\begin{tabular}{@{}lrrrrrr@{}}
\toprule
\textbf{State} & \textbf{Non} & \textbf{Summons} & \textbf{Mediation} & \textbf{Jurisdiction} & \textbf{Multiplicity} & \textbf{Contested} \\
& \textbf{Appearance} & \textbf{Trial} & & \textbf{Issue} & &\\
\midrule
Andhra Pradesh & 1674 & 814 & 260 & 210 & 124 & 753 \\
Chandigarh & 701 & 408 & 204 & 278 & 53 & 106 \\
Delhi & 2062 & 1359 & 1026 & 1045 & 208 & 521 \\
Goa & 381 & 78 & 90 & 33 & 18 & 109 \\
Gujarat & 5125 & 391 & 402 & 3059 & 107 & 797 \\
Haryana & 5214 & 3376 & 1731 & 2109 & 540 & 599 \\
Himachal Pradesh & 864 & 351 & 506 & 299 & 33 & 113 \\
Karnataka & 5850 & 2002 & 1846 & 953 & 410 & 3615 \\
Maharashtra & 5933 & 963 & 2546 & 3831 & 135 & 1118 \\
Punjab & 5821 & 3134 & 2100 & 2281 & 382 & 552 \\
\bottomrule
\end{tabular}
\end{table}

{\color{red} Please convert Table 11 in \% format with the first column mentioning the total number of cases for each State.}

>>>>

% \graphicspath{{../DOC/SLIDES/IMAGES/}}

@@

India has a slow judiciary --- courts are clogged with large backlogs \autocite{moog1992delays, debroy2008justice, dutta2019modernise}. As of 8th December 2021, over forty-one million cases were pending across district courts \autocite{njdg2021}. It is estimated that judicial delays cost India around 1.5\% of its GDP annually \autocite{dey2016_cost}. Slow judiciaries also have adverse consequences on the structure and efficiency of markets and the quality of life of citizens \autocite{world2004world, chemin2007impact, rao2020institutional}. Therefore, it is hardly surprising that tackling judicial delay has become a top priority for Indian judges and policymakers.

@@

Given this increasing volume and persistent mention of \gls{ni} cases as a major cause of judicial delay, \textcite{sridhar2017_cheque} examined the behaviour of cheque dishonour cases in India by looking at the DAKSH database of 67,000 cases spread across Indian courts. In most cases, the authors found that resolution is delayed well beyond the statutorily prescribed timeline of 180 days. Using 50,000 case level data for eight States and two Union Territories, we also estimate the survival probability (using non-parametric Kaplan-Meier estimation) to understand the proportion of unresolved cases within the stipulated time.\footnote{The dataset used for this analysis is documented in section \ref{sec:data-description}.} Figure \ref{fig:stateSurvival} shows the survival probability (case ongoing) for the \gls{ni} cases for different States. As per our estimate, less than 20\% of the cases get disposed of within the prescribed timeline of 180 days. Also, the probability of case completion within 500 days is less than 50\% for most States.

\begin{figure}[!ht]
\centering
\caption{Survival probability of cases across States}\label{fig:stateSurvival}
\footnotesize
\includegraphics[width=\textwidth]{surv_states-1.png}
\textit{Note:} The X-axis shows the number of days that a case takes from the date of admission until the outcome. The Y-axis shows the probability of the case continuing up to a given number of days.
\end{figure}

Since cheque dishonour cases are a large burden on the judiciary, various committees, institutions, and government bodies have focused on expediting disposal. For example, the \textcite{lci2008_138}, by relying on newspaper reports concerning the proportion of cheque dishonour cases, recommended setting up Fast Track Magisterial Courts (for a detailed discussion, refer to \textcite{bhan2015_placing}). The Act was also subject to amendments in 2015 and 2018 to deal with the various procedural matters. Similarly, the Supreme Court has issued directions to lower courts and various financial institutions to take a more pragmatic and realistic approach for the speedy disposal of \gls{ni} cases.

In 2020, the Supreme Court of India took on board a suo-motu case concerning the “expeditious trial of cases under section 138 of the Negotiable Instruments Act 1881” \autocite{sc2020_138}. Among other directions, the court formed a \gls{coe}\footnote{Headed by Hon’ble Mr. Justice RC Chavan, former Judge of the Bombay High Court.} and appointed Amici Curiae\footnote{Mr. Sidharth Luthra (Sr Advocate) and Mr. K Parameshwar (Advocate).} to assist the court, to study processes to expedite the disposal of complaints under \S~138 of the \gls{ni}. In late 2020, the Amici submitted their recommendations. This included (i) increasing the use of pre-and post-summons mediation, (ii) expediting the service of summons to reduce absconsion, (iii) addressing the multiplicity of proceedings, etc. (for additional detail, refer to \autocite{amicus2020_submission}).

Given this background, the purpose of this paper is fourfold. First, using case level eCourts data, for the first time in India, we estimate the real judicial caseload due to the \gls{ni}. Second, we estimate the average case disposal time and its variation over time and across States. Third, we build a conceptual framework to establish the link between specific case characteristics and procedural delay. Furthermore, we empirically test our conceptual framework of procedural determinants of case disposal time using regression analysis. In line with \textcite{bielen2015}, using case level interim and final orders data, we establish the link between the specific case characteristics and procedural issues while controlling for inherent court and judge characteristics. The empirical literature on the length of court proceedings currently focuses mainly on determinants at the country and court levels, e.g. public budgets for courts, backlogs, number of judges, etc. Some studies also focus on the case level determinants like the number of parties involved, number of pleadings, etc. Given access to the interim and final orders data, we focus on the case characteristics that can be directly targeted using procedural reforms.

@@

Our estimates suggest that cheque dishonour cases represent 13.2\% of courts' workload (pending and disposed) in India. This number could be underestimated because, in about 36\% of the cases in the sample, orders were either unavailable or could not be parsed. We also find significant variation in case disposal time across states. Median disposal time varies from as low as 218 days in Karnataka to as high as 547 days in Himachal Pradesh, with a median disposal time of 395 days for the dataset as a whole. We also find that only a fifth of the cases get disposed within the prescribed limit of six months.

Controlling for State and year fixed effects, our findings suggest that, all else being equal, cases where the accused fails to appear take an additional 201 days and 7 hearings to dispose of. Similarly, cases run as summons trials take 112 days and 7.4 hearings more than summary trials. Cases with jurisdictional issues take 271 days and 5.5 hearings more to dispose of than cases without these issues. Cases with a multiplicity of proceedings --- i.e. where multiple cheques are dishonoured and hence multiple, simultaneous cases are filed --- take an additional 168 days and 9.9 hearings to dispose of than those without. Contested cases take fewer days to dispose of than uncontested cases, even though they take more hearings. Our results also indicate that the 2015 amendments have significantly reduced case duration -- reducing the case duration by more than six months.

@@

\begin{enumerate}
\item how to bring the accused to the court?;
\item how to run the trials (summary or summons trials)?;
\item whether to refer the case to mediation;
\item how to clarify jurisdictional issues?; and
\item how to avoid the multiplicity of proceedings?
\item set up more courts.
\end{enumerate}

@@

We first drew a random sample of 100,000 cases filed between 1st January 2014 and 31st December 2018 across India from the database published by Development Data Lab.\footcite{devdatalabs2021_eCourtsData} Development Data Lab has made public a dataset of nearly all cases in the subordinate judiciary in India from 2010 to 2018, as reported by eCourts. This dataset contains the main reported fields for each case, such as state, district, act name, date of filing, date of disposal etc. This allowed us to pick a random sample from the universe of available data. We chose the time window --- 2014 to 2018 --- based on consultations with other researchers and practising advocates on the quality and completeness of data. We then attempted to download the case details for these 100,000 cases from the eCourts database.\footcite{ecourts2022} We were able to download case details for approximately 89,000 cases.\footnote{This difference was caused by non-functional links and corrupted or non-machine data.} We used regular expressions to check whether the ``Act Name'' field references the \gls{ni}. This was the first level of case classification. However, from our previous experience with eCourts data, we expected NI Act cases among those where the Act Name field did not match the search parameters.\footcite[The Act Name field sometimes does not contain the name of the substantive Act but rather the name of the procedural Act (e.g., CrPC) or, in rare cases, the subject matter of the case. This is mainly because case classification and nomenclature practices vary from State to State. However, in rare cases, the field is not populated or contains erroneous information. For details, please see,][]{damle2020_ecourtsData} So we used regular expressions to search for references to the \gls{ni}, cheque bounce/dishonour along with \S~138 in the interim and final order texts, where available. We tagged the cases found through this protocol as related to cheque dishonour under \S~138 of the \gls{ni}.

@@

\item [Contested cases] -- Whether or not a case was contested is reported by eCourts as part of the disposal type field. For example, a disposal type may be reported as ``contested-disposed'', ``uncontested-dismissed'' etc. Table \ref{tab:cont_yearWise} shows the number of cases in the sample which were contested by the accused. The proportion of such cases is significant but appears to have declined over the years. This may be due to an increased tendency of parties to compound a case.\footnote{\S~147 of the Negotiable Instruments Act provides that every offence punishable under the Act shall be compoundable. \S~147, Negotiable Instruments Act. Offences to be compoundable.} We expected contested cases to take more days and more hearings to dispose of than uncontested cases.

@@

\begin{table}[!ht]
\caption{Cases with identified characteristics of interest}
\label{tab:case_chars}
\centering
\footnotesize
\begin{tabular}{@{}lrrrrrrr@{}}
\toprule
\multirow{2}{*}{\textbf{State}} & \textbf{Total}& \textbf{Non} & \textbf{Summons} & \textbf{Mediation} & \textbf{Jurisdiction} & \textbf{Multiplicity} & \textbf{Contested} \\
& \textbf{Cases} & \textbf{Appearance} & \textbf{Trial} & & \textbf{Issue} & &\\
\midrule
  Andhra Pradesh & 2638 & 63.5 & 30.9 & 9.9 & 8.0 & 4.7 & 28.5\\
  Chandigarh & 731 & 95.9 & 55.8 & 27.9 & 38.0 & 7.3 & 14.5\\
  Delhi & 5202 & 39.6 & 26.1 & 19.7 & 20.1 & 4.0 & 10.0\\
  Goa & 399 & 95.5 & 19.5 & 22.6 & 8.3 & 4.5 & 27.3\\
  Gujarat & 6752 & 75.9 & 5.8 & 6.0 & 45.3 & 1.6 & 11.8\\
  Haryana & 5319 & 98.0 & 63.5 & 32.5 & 39.7 & 10.2 & 11.3\\
  Himachal Pradesh & 1164 & 74.2 & 30.2 & 43.5 & 25.7 & 2.8 & 9.7\\
  Karnataka & 11184 & 52.3 & 17.9 & 16.5 & 8.5 & 3.7 & 32.3\\
  Maharashtra & 8875 & 66.9 & 10.9 & 28.7 & 43.2 & 1.5 & 12.6\\
  Punjab & 5885 & 98.9 & 53.3 & 35.7 & 38.8 & 6.5 & 9.4\\
  \bottomrule
\end{tabular}
\end{table}
